RED WINE QUALITY EDA PROJECT by David Gonsalves
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

# Global code to supress unnecessary R code outputs
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
#install.packages("ggplot2", dependencies = T) 
#install.packages("knitr", dependencies = T)
#install.packages("dplyr", dependencies = T)
#install.packages("gridExtra", dependencies = T)
#install.packages("GGally", dependencies = T)

library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wineQR <- read.csv('wineQualityReds.csv')

# Examination of dataset variables
str(wineQR)
summary(wineQR)

# Based on the dataset description the Quality variable should be a factor
wineQR$quality <- factor(wineQR$quality, ordered = T)
# Focused xxamination of the Quality variable
str(wineQR$quality)
summary(wineQR$quality)
```

# EDA pertaining to Red Wine Quality.
The winQR dataset consists of tested chemical & physical properties (inputs) and
sensory (the output) variables related to red and white variants of the
Portuguese "Vinho Verde" wine. A total of 1,599 wines were tested.
Note: there is no data about grape types, wine brand, wine selling price, etc.
The inputs include objective tests (e.g. PH values) and the output is based on
sensory data (median of evaluations made by three wine experts). Each expert
graded the wine quality between 0 (very bad) and 10 (very excellent).
Description of variables:
Input variables (based on tested chemical and physical properties):
  1 - fixed acidity (tartaric acid - g / dm^3)
  2 - volatile acidity (acetic acid - g / dm^3)
  3 - citric acid (g / dm^3)
  4 - residual sugar (g / dm^3)
  5 - chlorides (sodium chloride - g / dm^3
  6 - free sulfur dioxide (mg / dm^3)
  7 - total sulfur dioxide (mg / dm^3)
  8 - density (g / cm^3)
  9 - pH
  10 - sulphates (potassium sulphate - g / dm^3)
  11 - alcohol (% by volume)
Output variable (based on sensory data): 
  12 - quality (score between 0 and 10)

Dataset source:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. [2009]
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots1}

create_plot <- function(varname, binwidth = NULL, xlab = "") {
  return(qplot(x=wineQR[varname], data=wineQR,                                              binwidth=binwidth, xlab=xlab, fill=I("red")))
}

#User defined function to plot preliminary histograms

grid.arrange(create_plot('fixed.acidity', 0.1,'fixed.acidity'),
             create_plot('volatile.acidity', 0.01,'volatile.acidity'),
             create_plot('citric.acid', 0.01,'citric.acid'),
             create_plot('residual.sugar', 0.1,'residual.sugar'),
             create_plot('chlorides', 0.001,'chlorides'),
             create_plot('free.sulfur.dioxide', 1,'free.sulfur.dioxide'),
             create_plot('total.sulfur.dioxide', 1,'total.sulfur.dioxide'),
             create_plot('density', 0.0001,'density'),
             create_plot('pH', 0.01,'pH'),
             create_plot('sulphates', 0.01, 'sulphates'),
             create_plot('alcohol', 0.1,'alcohol'),
             #create_plot('quality', 0.01,'quality'),
             qplot(wineQR$quality, fill=I("red"), xlab='quality'),
             ncol = 4)

# Preliminary histogram plots to explore the shapes of the data distributions

```

The following data pattern for each chemical, physical, or sensory property
were observed:

* Normally distributed properties: density; PH; quality
* Long tailed properties: fixed acidity; volatile acidity; free sulfur; total
sulfur; alcohol
* Bimodal property: citric acid
* Properties with significant outliers: residual sugar; chlorides; sulphates

```{r echo=FALSE, Univariate_Plots2}

fa_log10 <- qplot(data = wineQR, x = log10(fixed.acidity))

tsd_log10 <- qplot(data = wineQR, x = log10(total.sulfur.dioxide))

ca_log10 <- qplot(data = wineQR, x = log10(citric.acid))

alc_log10 <- qplot(data = wineQR, x = log10(alcohol))

rs_log10 <- qplot(data = wineQR, x = log10(residual.sugar))

chl_log10 <- qplot(data = wineQR, x = log10(chlorides))

sulph_log10 <- qplot(data = wineQR, x = log10(sulphates))

# Plots transforming long tailed and bimodal data to uncover normal 
# distributions

grid.arrange(fa_log10, ca_log10, tsd_log10, alc_log10, rs_log10, chl_log10,               sulph_log10,  ncol = 3)

# Combined display of transformed data

```

The summary(wineQR$quality) results indicate a potential convenience for
introducing a new variable to combine the wine quality into groups,
as follows: low (quality 3 or 4), medium (5 or 6), and high (7 or 8)

```{r echo=FALSE, Univariate_Plots3}

wineQR$Q.group <- NA
wineQR$Q.group <- ifelse(wineQR$quality < 5, 'low (3-4)',
                  ifelse(wineQR$quality < 7, 'medium (5-6)',
                  'high (7-8)'))
wineQR$Q.group <- factor(wineQR$Q.group, levels = c('low (3-4)',
                          'medium (5-6)', 'high (7-8)'))

# Created a new variable grouping wine quality into three categories

summary(wineQR$Q.group)

# Verifying wine quality groupings

qplot(x=wineQR$Q.group, data = wineQR, fill=I("red"), xlab='Quality Groups')

# Histogram illustrating the effect of adding variable Q.group to the
# dataset 
```

# Univariate Analysis

Questions for Relection:

### What is the structure of the dataset?
After conversion from csv the structure of the wineQR dataset is a dataframe in
wide format. An additional variable was added to group the quality data.

### What is/are the main feature(s) of interest in the dataset?
The main features of the dataset are:
1. All the wines are red and were produced as wine type "vinho verde" which
indicates the wines originate from a particular region in Portugal, are not
aged, and exhibit a slight effervescence.
2. The data for each wine contains 10 chemical properties, one physical property (density), and one sensory property (quality).
3. My main interest is to investigate Which chemical and physical properties
influence the quality of this type of red wine?

### What other features in the dataset will help support the investigation?
The sensory score for each wine is the median of at least 3 evaluations made by
wine experts, a traditional way of evaluating wines. This feature provides a
subjective, consensus measure to lend some credence to any findings.

### Did you create any new variables from existing variables in the dataset?
To simplfiy the analysis I introduced a new variable grouping the quality data
into Q.group with values low (3-4), medium (5-6), and high (7-8).

### Of the features you investigated, were there any unusual distributions?
The possibly unusual (unexpected) distribution was 
Did you perform any operations on the data to tidy, adjust, or change the form
of the data? If so, why did you do this?
The data was clean so no tidying was needed. When I reviewed the log
transformations it seemed that citric acid, sulphates, and alcohol have the most potential to be wine quality differentiators.


# Bivariate Plots Section

The Univariate analysis indicate some wine properties of interest namely,
citric acid, sulphates, an alcohol but did not illuminate any relationships
between properties. My next step was to employ the ggpairs function to pair sets
of variables in the wineQR dataset in order to uncover possible
interrelationships between the properties.

```{r echo=FALSE, Bivariate_Plots1, fig.width = 14.5, fig.height = 14.5}

library(GGally)
wineQR.subset <- wineQR[ , !names(wineQR) %in% c("X","Q.group")]

# Subset the wineQR dataset to allow ggpairs plot to be more efficient

ggpairs(wineQR.subset,
  lower = list(continuous = wrap("points", shape = I('.'))),
  upper = list(combo = wrap("box", outlier.shape = I('.'))))

#Pair up wineQR dataset properties to look for relationships

```

The ggpairs plot indicates that correlations between many of the wine
properties are weak.
Fixed acidity and volatile acidity seem to be somewhat inversely correlated.
Clearly pH is expected to be inversely related with acidity since pH is a
measure of acidity [1(strong) - 6(weak)], distilled water [7(neutral)], basicity [8(weak) - 14(strong)].
Although there is separate data for fixed, volatile, and citric acids the
correlation data, and common sense, indicate that these measures probably
overlap.
There also seems to be a positive correlation between density and acidity. Since
density is a physical property and acidity a chemical property intuitively more
wine molecules would carry more acid components.

In summary, ggpairs boxplots seem to be more useful in terms of addressing the
central question regaring how the various properties influence the perceived
quality of the wines. Therefore I focused on studying the boxplots more closely.


```{r echo=FALSE, Bivariate_Plots2, fig.width = 8.5}

grid.arrange(qplot(data = wineQR, x = Q.group, y = fixed.acidity,
                    geom = 'boxplot'), 
             qplot(data = wineQR, x = Q.group, y = volatile.acidity,
                    geom = 'boxplot'),
             qplot(data = wineQR, x = Q.group, y = citric.acid,
                    geom = 'boxplot'),
             ncol = 3)

# Searching for patterns between acidic properties, by quality group

grid.arrange(qplot(data = wineQR, x = Q.group, y = free.sulfur.dioxide,
                   geom = 'boxplot'),
             qplot(data = wineQR, x = Q.group, y = total.sulfur.dioxide,
                   geom = 'boxplot'),
             ncol = 2)

# Searching for patterns between free and total sulphur dioxide, by quality
# group

grid.arrange(qplot(data = wineQR, x = Q.group, y = residual.sugar,
                    geom = 'boxplot'),
             qplot(data = wineQR, x = Q.group, y = chlorides,
                    geom = 'boxplot'),
             ncol = 2)

# Investigating relative significance of sugars and chlorides, by quality group

grid.arrange(qplot(data = wineQR, x = Q.group, y = density,
                   geom = 'boxplot'),
             qplot(data = wineQR, x = Q.group, y = sulphates,
                   geom = 'boxplot'),
             ncol = 2)

# Investigating relative significance of density and sulphates, by quality group

qplot(data = wineQR, x = Q.group, y = alcohol, geom = 'boxplot')

# Investigating relative significance of alcohol content, by quality group

```

I confirmed with these plots that measures of fixed acidity and volatile
acidity seem to counterbalance each other. On the other hand, citric acid seems
to play a more signifiant role in distinguising the higher quality wines. To
further condense my findings, I've decided to combine the fixed and volatile
acidities together in a new variable labeled FaV.acidity.

```{r echo=FALSE, Bivariate_Plots3}

wineQR$FaV.acidity <- wineQR$fixed.acidity + wineQR$volatile.acidity

# Combined Fixed and Volatile acidity into a new single variable FaV.acidity

grid.arrange(qplot(data = wineQR, x = Q.group, y = FaV.acidity,
                    geom = 'boxplot'), 
             qplot(data = wineQR, x = Q.group, y = citric.acid,
                    geom = 'boxplot'),
                    ncol = 2)

# Plot of combined acidity alongside citric acid to better guage which of these
# properties is likely to be more inluential in distiguishing the higher
# quality wines in our dataset.

```

The boxplot investigations strengthend my expectation that citric acid, alcohol,
and sulphates are likely good indicators for wine quality in this dataset.
Consequently, I obtained quantitative summaries for these properties by quality
group.


```{r echo=FALSE, Bivariate_Plots4}

with(wineQR, by(citric.acid, Q.group, mean))
with(wineQR, by(alcohol, Q.group, mean))
with(wineQR, by(sulphates, Q.group, mean))

# Obtain the means for each propery of interested by quality group

```

# Bivariate Analysis

The boxplots strengthen my expectation that citric acid, alcohol, and
sulphates are likely good indicators for wine quality in this dataset.
Consequently, I obtained quantitative summaries for these properties by quality
group

### What are relationships I observed in this part of the investigation?
From the correlation coefficients in the ggpairs plot and from the subsequent
boxplots it has become clear that there is a significant positive relationship
between citric acid concentration and quality; between between alcohol content
and quality; and between sulphates concentration and quality These relationships
have very few outliers within the high quality group.

### What was the strongest relationships you found?
Comparing the quantitative summary statistics by quality group I see that for
citric acid the proportional difference in means between medium and high
qualities is (0.3765 - 0.2583)/0.2585 = 45.8%
For alcohol acid the proportional difference in means between medium and high
qualities is (11.52 - 10.25)/10.25 = 12.4%
For sulphates the proportional difference in means between medium and high
qualities is (0.7435 - 0.6473)/0.6473 = 14.9%

I conclude that citric acid is the most distinguishing factor in determining
the "vinho verde" wine quality, followed by the percentage of sulphates, and
then the concentration of alcohol
Revisiting the ggpairs plot I note that the correlation coefficient between
citric acid and sulphates is 0.313. So there does appears to be a weak
relationship between these two properties.

### Did you observe any interesting relationships between the other features?
Wine properities other than citric acid, sulphates, and alcohol do not appear to
be promising predictors of wine quality.

# Multivariate Plots Section

Given the findings above, I want to explore relationships between the
properties that are the most promising for predicting wine quality.

```{r echo=FALSE, Multivariate_Plots1}

ggplot(data = wineQR, aes(x = citric.acid, y = sulphates)) +
  geom_point(aes(color = Q.group), stat = 'summary', fun.y = median) +
  geom_smooth(method = "lm", se = TRUE, size = 1) +
  guides(colour = guide_legend(reverse=T))

# Scatter plot showing how sulphate concentrations vary with increased citric
# acid concentrations by wine quality group
```

```{r echo=FALSE, Multivariate_Plots2}

ggplot(data = wineQR, aes(x = citric.acid, y = alcohol)) +
  geom_point(aes(color = Q.group), stat = 'summary', fun.y = median) +
  geom_smooth(method = "lm", se = TRUE, size = 1) +
  guides(colour = guide_legend(reverse=T))

# Scatter plot showing how alcohol content varies with increased citric acid concentrations by wine quality group

```

# Multivariate Analysis

Were there any interesting or surprising interactions between features?
For all three groups of wine quality, the variations around the mean for alcohol
content were quite pronounced. By contrast, the variations around the mean for
sulphates were small. It appears that citric acid together with sulphate
concentrations may be the reliable predictors of wine quality for this dataset.
The standard error shading reiforces this opinion.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

wQR_b1 <- ggplot(data = wineQR, aes(x = quality, fill=I("red"),                             col=I("black"))) +
          geom_bar(stat = "count") +
          xlab("Quality Rating") + ylab("Count") +
          ggtitle("Vinho Verde Red Wines by Quality")
  
wQR_b2 <- ggplot(data = wineQR, aes(x = Q.group, fill=I("red"),                             col=I("black"))) +
          geom_bar(stat = "count") +
          xlab("Quality Group") + ylab("Count") +
          ggtitle("Vinho Verde Red Wines by Quality Group")

grid.arrange(wQR_b1, wQR_b2,ncol = 2)

# Side-by-side bar plots of quality ratings - discrete and grouped

```

### Description One
My main interest for this EDA project is to investigate Which chemical and
physical properties influence the quality of "vinho verde" red wines. The key
starting point is to understand how the wine quality was measured. Evaluations
were made by three wine experts. Each expert graded the wine quality between
0 (very bad) and 10 (very excellent). These bar charts show the counts for each
quality grade in the dataset containing a total of 1,599 red wines. The second
bar chart summarizes the quality into three groups namely, low (3-4),
medium (5-6), and high (7-8). This grouping helps to simplify the analysis and
the vizualizations.

### Plot Two
```{r echo=FALSE, Plot_Two, fig.width = 8.5}

grid.arrange(qplot(data = wineQR, x = Q.group, y = citric.acid,                                 geom=c("boxplot"), fill=Q.group,
                    main="Citric Acid by Group",
                    xlab="", ylab="Citric Acid (g/cubic dm)") +
                    theme(legend.position="none"),
            qplot(data = wineQR, x = Q.group, y = sulphates,                                    geom=c("boxplot"), fill=Q.group,
                    main="Sulphates by Group",
                    xlab="", ylab="Sulphates (g/cubic dm)") +
                    theme(legend.position="none"),
            qplot(data = wineQR, x = Q.group, y = alcohol,                                      geom=c("boxplot"), fill=Q.group,
                    main="Alcohol by Group",
                    xlab="", ylab="Alcohol (% by volume)") +
                    theme(legend.position="none"),
            ncol = 3)

# Side-by-side boxplots illustrating the relative impact of each of these
# wine properties on the three groups of wine quality

```

### Description Two
These plots clearly illustrate the significance of citrus, sulphates and
alcohol as differentiators between the low, medium and high quality wines. When
analyzing the boxplots it is important to note the differences between the
y-scales. The quantitative analysis described earlier shows that the differences
between the means of medium and high quality wines are 46% for citric acid; 15%
for sulphates; and 12% for alcohol.

### Plot Three

```{r echo=FALSE, Plot_Three, , fig.width = 10, fig.asp = 1}

wQR_m1 <- ggplot(data = wineQR, aes(y = citric.acid, x = sulphates, color =               Q.group)) +
              geom_point(alpha = 0.8, size = 1) +
              geom_smooth(method = "lm", se = TRUE, size = 1) +                           scale_y_continuous(limits = c(0,0.8)) +                                     scale_x_continuous(limits = c(0.1,2)) +
              ylab("Citric Acid (g/cubic dm)") +
              xlab("Sulphates (g/cubic dm)") +
              scale_color_brewer(type = 'seq', palette = 11,                                                  guide = guide_legend(title =
              'Quality Groups', reverse = TRUE)) +
              theme(legend.position="bottom") +
              ggtitle("Citric Acid vs Sulphates over Wine Quality Groups")

wQR_m2 <- ggplot(data = wineQR, aes(y = citric.acid, x = alcohol,
              color = Q.group)) +
              geom_point(alpha = 0.8, size = 1) +
              geom_smooth(method = "lm", se = TRUE, size = 1) +
              scale_y_continuous(limits = c(0,0.8)) +
              ylab("Citric Acid (g/cubic dm)") +
              xlab("Alcohol (% by volume)") +
              scale_color_brewer(type = 'seq', palette = 11,                                                  guide = guide_legend(title =
                                  'Quality Groups', reverse = TRUE)) +
              theme(legend.position = "bottom") +
              ggtitle("Citric Acid vs Alcohol over Wine Quality Groups")

grid.arrange(wQR_m1, wQR_m2,ncol = 2)

# Side-by-side multivariate plots showing the relationships between citric acid
# concentrations and sulphate concentrations and between citric acid
# concentrations and alcohol concentrations juxtaposed with a dot plot of the
# wine quality groups
```

### Description Three
Plot three is similar to the plots in the multivariate plots section. The
difference is that the properties are flipped on the axes, and the linear model
smoothing is applied to each quality group. 
Clearly, plot three shows that the high quality wines in this dataset tend to
be found in the upper two quadrants of each plot. Also, the shapes of the
confidence level shadows indicate that the linear model is not likely to be a
reliable predictor of the relationships between these properties. In other words
these three properties seem to be quite independent of each other.  

------

# Reflection

I've long been facinated by wine making - in particular what makes a wine good
This dataset of 1,599 "vinho verde" red wines, containing 10 objective measures
of important wine properties as well as expert wine quality rankings, gave me
the opportunity to explore this question - using EDA techniques.

I think one good data analyst decision that helped me perform the analysis more
easily and that enables the reader to better understand the results is to group
the quality ratings into low (3-4), medium (5-6), and high (7-8).
I've now a greater appreciation for the chemistry of wines and I've determined
that three properties citric acid, sulphates, and alcohol are important
components in the best "vinho verde"" wines.

However, coming up with some formula that could be used to predict, or to
produce, a high quality "vinho verde" red wine has proven to be illusive, so far
This begs the question, would a much bigger and wider dataset allow a researcher
to draw more predictive conclusions? For example, California is renowned for
consistently producing good wines, but in certain years similar European wines
far surpass Californian wines in quality. Why? Experts attribute this difference
to the relative consistency of California's climate versus the more erratic
European climate. So, weather conditions in a given season, soil composition,
and grape varietals are also important factors in determining wine quality. Data
on these factors should contribute to a more predictive model for wine quality.

My analysis, coupled with more general information about wine making, leads me
to the view that the quality of wine, a product honed over thousands of years
and evaluated by diverse palates, most likely cannot be reduced to any single
scientific equation. 
